Numerical simulations of the Cordilleran ice sheet through the last glacial cycle

### 2014-07-17 cordillera-cycle.tex
Polish 4.1 (parts missing).

% -----------------------------------------------------------------------------
\subsection{Overview}
% -----------------------------------------------------------------------------

@@

% =============================================================================
\section[Comparison to the geologic record]
        {Comparison to the geologic record\renote{
    Arjen, you have proposed 2.1 Evaluation of model runs agains field
    evidence, 2.2 Palaeoglaciology of the Cordilleran ice sheet, and 2.3 The
    last deglaciation; but I don't think these headers work. In my vision of
    the paper, this entire ``discussion'' part is all about evaluation of model
    runs agains field evidence and palaeoglaciology. Because of this, and
    because we also present new ``results'', or rather different views of the
    same results, I propose to rename it to a more explicit ``Comparison to the
    geologic record'' (or similar). I have also renamed the ``results'' to
    ``Sensitivity to atmospheric forcing time-series'', which is actually what
    that section is about: raw model output, no discussion against the geologic
    record.}}
\label{sec:discussion}
% =============================================================================

@@

Independently of the choice of palaeo-temperature record\renote{
    In my view, ``atmospheric/climate forcing'' would be far too broad here. We
    could possibly apply a much wilder choice of atmospheric forcing
    techniques, leading just as wild differences in model output
    \citep[e.g.,][but possibly much more than that]{Seguinot.etal.2014}.}
used to force the ice sheet model, our simulations consistently produce two
glacial maxima during the last glacial cycle. The first maximum configuration
is obtained during MIS~4 (61.9--55.4\,ka) and the second during MIS~2
(29.5--16.9\,ka). These events broadly correspond in timing to the Gladstone
(MIS~4) and McConnell (MIS~2) glaciations documented by geological evidence for
the northern sector of the Cordilleran ice sheet
    \citep{Duk-Rodkin.etal.1996, Ward.etal.2007,
           Stroeven.etal.2010, Stroeven.etal.2014},
and to the Fraser Glaciation (MIS~2) documented for its southern sector
    \citep{Porter.Swanson.1998, Margold.etal.2014}.
There is patchy stratigraphical evidence for glaciations older than the Fraser
Glaciation \citep{Clague.Ward.2011} in British Columbia, and their extent and
timing are therefore still highly conjectural
    \citep[perhaps MIS~4 or early MIS~3; e.g.,][]{Cosma.etal.2008}.

However, the exact timing of modelled MIS~2 maxima depend strongly on the
choice of applied palaeo-temperature record, which allows for a more in-depth
comparison with geological evidence for the timing of maximum Cordilleran ice
sheet extent. In the Puget lowland (Fig.~\ref{fig:locmap}), the LGM advance of
the southern Cordilleran ice sheet margin has been constrained by radiocarbon
dating on wood between 17.4 and 16.4\,\unit{\chem{^{14}C}\,cal\,kyr\,BP}
\citep{Porter.Swanson.1998}.
These dates have been confirmed by the offshore sediment record, which shows an
increase of glaciomarine sedimentation between 19.5 and
16.2\,\unit{\chem{^{14}C}\,cal\,kyr\,BP} \citep{Cosma.etal.2008}. Radiocarbon
dating of the northern Cordilleran ice sheet margin is much less constrained
but straddles presented constraints from the southern margin. However,
cosmogenic exposure dating by \citep{Stroeven.etal.2010, Stroeven.etal.2014}
places the timing of maximum CIS extent during the McConnell glaciation close
to 17\,ka.

Among the simulations presented here, only those forced with the GRIP, EPICA
and Vostok palaeo-temperature records yield Cordilleran ice sheet maximum
extent that may be compatible with these field constraints
(Fig.~\ref{fig:timeseries}, lower panel; Fig.~\ref{fig:snapshots}, lowest
panels). Simulations driven by the NGRIP, ODP~1012 and ODP~1020
palaeo-temperature records, on the contrary, yield MIS~2 maximum Cordilleran
ice sheet extents that pre-date field-based constraints by several thousands of
years. Concerning the simulations driven by oceanic records, this early
deglaciation is caused by an early warming present in the alkenone
palaeo-temperature reconstructions (Fig.~\ref{fig:timeseries}, upper panel;
\citealp[Fig.3]{Herbert.etal.2001}). However, this
early warming is a local effect, corresponding to a weakening of the California
current \citep[Fig.~3]{Herbert.etal.2001}. The California current, driving cold
waters southwards along the south-western coast of North America\renote{
    Not only USA but also Baja California in Mexico.},
has been shown to have weakened during each peak global glaciation (in SPECMAP)
durint the past 550\,ka, resulting in paradoxically warmer sea-surface
temperatures at the locations of the ODP~1012 and ODP~1020 sites also during
the LGM \citep{Herbert.etal.2001}.

@@

\todo{say something about NGRIP.}

... Hence, we focus the rest of our analysis on simulations forced by palaeo-temperature records from GRIP and EPICA ice cores (Fig.~\ref{fig:timeseries}, lower panel, dotted lines).

\todo{avoid repetition with the next paragraph.}

These results indicate that a palaeo-temperature proxy record located in the
direct vicinity of the former Cordilleran ice sheet would potentially lead to
improved understanding of the history of advance and retreat of the western ice
sheet margin. Because, to our knowledge, no
such record yet exists, we focus the rest of our analysis on simulations forced
by palaeo-temperature records from GRIP and EPICA ice cores
(Fig.~\ref{fig:timeseries}, lower panel, dotted lines).

@@

\todo{Add MIS~2 ice topo and velocity map. Include two paragraphs from Martin.}

@@

\todo{Add MIS~4 ice topo and velocity map. Discuss that some of the margins
      are more extensive during MIS~4 than during MIS~2, in the GRIP simu.}

>>>>

% -----------------------------------------------------------------------------
\subsection{Overview}
\label{sec:overview}
% -----------------------------------------------------------------------------

@@

% =============================================================================
\section[Comparison to the geologic record]
        {Comparison to the geologic record\renote{
    Arjen has proposed: 2.1 Evaluation of model runs agains field evidence, 2.2
    Palaeoglaciology of the Cordilleran ice sheet, and 2.3 The last
    deglaciation; but I don't think that these headers work. In my view, this
    entire ``discussion'' part is all about evaluation of model runs agains
    field evidence and palaeoglaciology. It is necessary that this comparison
    starts here, in order to ``kick out'' the records that perform badly (ODP*,
    NGRIP) and justify that the rest of the analysis is performed on two
    (high-resolution) simulations only. Such figures as
    Figs.~\ref{fig:duration}--\ref{fig:erosion} and Fig.~\ref{fig:deglac},
    which I consider as different views of the results rather than new results
    (apart the fact that they use higher resolution) are drawn for the purpose
    of comparing model results against geological evidence. I propose to rename
    this section to a more explicit ``Comparison to the geologic record'' (or
    similar). I have also renamed the ``results'' to ``Sensitivity to
    atmospheric forcing time-series'', which is actually what that section is
    about: raw model output, no discussion against the geologic record. This
    new discussion is divided in three parts: 4.1 Glacial maxima, where we
    study the maximum stages, relatively well-documented by the geology; 4.2
    Nucleation centres, where we study ``intermediate'' states, for which less
    is known, but the model can say something; and 4.3 The last deglaciation,
    which deserves a separate section because many people are interested in
    this period, and much geological evidence is available.}}
\label{sec:discussion}
% =============================================================================

@@

Independently of the choice of palaeo-temperature record\renote{
    In my view, ``atmospheric/climate forcing'' would be far too broad here. We
    could possibly apply a much wilder choice of atmospheric forcing
    techniques, leading just as wild differences in model output
    \citep[e.g.,][but possibly much more than that]{Seguinot.etal.2014}.}
used to force the ice sheet model, our simulations consistently produce two
glacial maxima during the last glacial cycle. The first maximum configuration
is obtained during MIS~4 (61.9--55.4\,ka) and the second during MIS~2
(29.5--16.9\,ka; Figs.~\ref{fig:timeseries} and~\ref{fig:snapshots};
Table~\ref{tab:extrema}). These events broadly correspond in timing to the
Gladstone
(MIS~4) and McConnell (MIS~2) glaciations documented by geological evidence for
the northern sector of the Cordilleran ice sheet
    \citep{Duk-Rodkin.etal.1996, Ward.etal.2007,
           Stroeven.etal.2010, Stroeven.etal.2014},
and to the Fraser Glaciation (MIS~2) documented for its southern sector
    \citep{Porter.Swanson.1998, Margold.etal.2014}.
There is patchy stratigraphical evidence for glaciations older than the Fraser
Glaciation \citep{Clague.Ward.2011} in British Columbia, and their extent and
timing are therefore still highly conjectural
    \citep[perhaps MIS~4 or early MIS~3; e.g.,][]{Cosma.etal.2008}.

However, the exact timing of modelled MIS~2 maxima depends strongly on the
choice of applied palaeo-temperature record, which allows for a more in-depth
comparison with geological evidence for the timing of maximum Cordilleran ice
sheet extent. In the Puget lowland (Fig.~\ref{fig:locmap}), the LGM advance of
the southern Cordilleran ice sheet margin has been constrained by radiocarbon
dating on wood between 17.4 and 16.4\,\unit{\chem{^{14}C}\,cal\,ka\,BP}
\citep{Porter.Swanson.1998}.
These dates have been confirmed by the offshore sedimentary record, which shows
an increase of glaciomarine sedimentation between 19.5 and
16.2\,\unit{\chem{^{14}C}\,cal\,ka\,BP} \citep{Cosma.etal.2008}. Radiocarbon
dating of the northern Cordilleran ice sheet margin is much less constrained
but straddles presented constraints from the southern margin. However,
cosmogenic exposure dating
places the timing of maximum CIS extent during the McConnell glaciation close
to 17\,ka \citep{Stroeven.etal.2010, Stroeven.etal.2014}.

Among the simulations presented here, only those forced with the GRIP, EPICA
and Vostok palaeo-temperature records yield Cordilleran ice sheet maximum
extent that may be compatible with these field constraints
(Fig.~\ref{fig:timeseries}, lower panel; Table~\ref{tab:extrema}).
Simulations driven by the NGRIP, ODP~1012 and ODP~1020
palaeo-temperature records, on the contrary, yield MIS~2 maximum Cordilleran
ice sheet extents that pre-date field-based constraints by several thousands of
years. Concerning the simulations driven by oceanic records, this early
deglaciation is caused by an early warming present in the alkenone
palaeo-temperature reconstructions (Fig.~\ref{fig:timeseries}, upper panel;
\citealp[Fig.~3]{Herbert.etal.2001}). However, this
early warming is a local effect, corresponding to a weakening of the California
current \citep{Herbert.etal.2001}. The California current, driving cold
waters southwards along the south-western coast of North America\renote{
    Not only USA but also Baja California in Mexico.},
has been shown to have weakened during each peak global glaciation (in SPECMAP)
durint the past 550\,ka, including the LGM, resulting in paradoxically warmer
sea-surface temperatures at the locations of the ODP~1012 and ODP~1020 sites
\citep{Herbert.etal.2001}.

@@

\todo{Write something about NGRIP.}

Hence, we focus the rest of our analysis on simulations forced by
palaeo-temperature records from GRIP and EPICA ice cores. To allow for a more
detailed comparison against the geologic record, these two simulations were
re-run on the high-resolution grid (Fig.~\ref{fig:timeseries}, lower panel,
dotted lines).

@@

\todo{Add MIS~2 ice topo and velocity map. Include two paragraphs from Martin.
      Describe model limitations referring to \citet{Seguinot.etal.2014}}

@@

\todo{Add MIS~4 ice topo and velocity map. Discuss that some of the margins
      are more extensive during MIS~4 than during MIS~2 in the GRIP simulation}
